---
title: "Imputation & Weighting example"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Imputation

This is an example code of missing data imputation using MICE package.

```{r cars}
## Import example data
data <- airquality
data_sub <- data[,c(1,2)]
summary(data_sub)

## Quick view of missing data
pMiss <- function(x){sum(is.na(x))/length(x)*100}
apply(data_sub, 2, pMiss)

## Look at missing data pattern
library(mice)
md.pattern(data_sub)
```

Results indicated that 111 observations have complete data, 35 observations have missing data in variable Ozone, 5 observations have missing data in variable Solar.R, and 2 observations have missing data on both variables.

```{r, include=TRUE,warning=FALSE,error=FALSE}
## Look at missing data pattern with more plots
library(VIM)
plot1 <- aggr(data_sub, col=c('blue','pink'), number=TRUE,
              sortVars=TRUE, labels=names(data_sub), cex.axis=.7,
              gap=3, ylab=c("Histogram of missing data",
                            "Pattern"))
```

The plots showed that 72.50% of data are complete, 22.90% of data has missing value on Solar.R variable, 3.30% of data has missing value on Ozone variable, and 1.30% of data has missing values on both variables.

```{r, include=TRUE,warning=FALSE,error=FALSE}
## Imputing missing data
# using predictive mean matching imputation
data_impute <- mice(data_sub, m=5, meth='pmm', maxit=50, seed=500)
summary(data_impute)

## Check the imputed data for variables
data_impute$imp$Ozone
data_impute$imp$Solar.R

## Get the completed dataset after imputing missing data
data_complete <- complete(data_impute, 1)
md.pattern(data_complete)
```

Results indicated that now there is no missing value in the dataset.

```{r, include=TRUE,warning=FALSE,error=FALSE}
## Inspecting the distribution of original and imputed data
densityplot(data_impute)
```

We expect the distributions of original and imputed dataset to be similar. The distribution of original dataset is plotted in blue and the imputed dataset is plotted in red. As the plots showed, for both variables, the distributions of original and imputed dataset are similar.

## Weighting

The following code of survey weights will be demonstrated using a fake dataset. This weighting process will be mimic the situation where we have certain demongraphic variables (gender in the following analysis) may not be representative of the population. The assumption is that the population variable is available.

```{r, include=TRUE,warning=FALSE,error=FALSE}
## Generate variables and created dataset
set.seed(12345)
gender = c("Female", "Male")
gender = sample(gender, 100, replace = TRUE)
gender = as.numeric(factor(gender))
ethnicity = c("White", "African_American", "Other")
ethnicity = sample(ethnicity, 100, replace = TRUE)
ethnicity = as.numeric(factor(ethnicity))
income = c(0:100000)
income = sample(income, 100, replace = TRUE)

data = cbind(gender,ethnicity,income)
data = as.data.frame(data)

## Create unweighted dataset with survey package
library(survey)
data.svy.unweighted <- svydesign(ids=~1, data=data)
```

Next, the weighting will be performed based on the population gender probabilities. I assume the population values for female (1) and male(2) are .45 and .55.

```{r, include=TRUE,warning=FALSE,error=FALSE}
gender.dist <- data.frame(gender = c("1","2"),
                          Freq =nrow(data)*c(0.45, 0.55))
```

rake function in survey package will be used to weight the data by population gender values.

```{r, include=TRUE,warning=FALSE,error=FALSE}
data.svy.rake <- rake(design = data.svy.unweighted,
                      sample.margins = list(~gender),
                      population.margins = list(gender.dist))
```

In case the weights is too large or too small, I put limits on the weights using the trimWeights function.

```{r, include=TRUE,warning=FALSE,error=FALSE}
data.svy.rake.trim <- trimWeights(data.svy.rake, 
                                  lower=0.3, upper=3,
                                  strict=TRUE)
```

Next, I'm going to compare mean of variables in the weighted dataset and the original dataset.

```{r, include=TRUE,warning=FALSE,error=FALSE}
svymean(data, data.svy.rake.trim) # weighted dataset
apply(data, 2, mean) # original dataset
```